\textbf{\sc Deliberative Autonomy for Marine Robotics}

### 2012-01-27 trunk/future.tex
update
updates

>>>>

% 1.6 Future Work?
% explain where EUROPA is going? (ANML, incorporation of classical planning algorithms, soft constraints, better search, etc).

% The main advantage of CP is the richness of the constraints that can be modeled and the explicit use of logical inference to prune the search space

The Ocean Sciences are at a cusp, straddling between traditional
ship-based exploration with newer observatory-based methods. In the
United States the NSF funded Ocean Observatories Initiative \cite{ooi}
and regional bodies such as \texttt{CeNCOOS} \cite{cencoos} are
investing in new observational methods and technologies which propose
to aid oceanographers by producing substantially higher-resolution
data. These will use moorings, cabled observatories and glider fleets
generating near real-time data from one or more locations as a
continuous stream. The tack chosen by \can has been to complement such
methods with the aim of developing methods for a \emph{portable
  mobile} observatory which is dependent on autonomous platforms. In
this context what we advocate in this chapter is an initial step in
making robotic platforms more adaptive, scaling them towards fleets of
vehicles (underwater or on the surface) and most importantly using
them to solve the larger challenge of Sampling spatio-temporal fields.

While our trajectory of research in the oceanographic domain follows
that in spacecraft autonomy, from onboard \cite{mus98} to
mixed-initiative methods \cite{bresina05}, the challenges are
significantly different with the environment playing a far larger
role. Moreover, prediction and projection of a future course of action
for a robot given the level of uncertainty calls for representational
and architectural methods which would enable adaptation at different
levels of abstraction, from mission-planning to low level actuation.

Our efforts were initially targeted to full robotic autonomy; while
this is an important goal that we continue to strive towards, it is
increasingly clear that human-in-the-loop methods also have a sizable
role to play in ocean exploration and by extension to Marine
Robotics. Not only is an embedded \rx an important element of the
effort towards smarter vehicles, but we are working towards an
implementation of \rx which will perform multi-vehicle plan synthesis
for the \od.

A key role that \rx has played and will continue to play is towards
\emph{event response} scenarios. In Section \ref{sec:results} we show
the role Machine Learning (ML) techniques have played thus far in
in-situ plan adaptation. This interaction between Planning and ML we
believe will play a significant role going forward in the general
theme of exploration and discovery. Interpretation of incoming sensor
data to deal with scientific surprise and readapt for opportunistic
science will be critical to dealing with undersampling the
ocean\footnote{Walter Munk has famously stated \emph{``Most of the
    previous century could be called a “century of undersampling”''}
  -- Testimony to the U.S. Commission On Ocean Policy, 18 April
  2002}. Much of what we know of the ocean was derived by exploring in
the blind. We anticipate longer duration robotic vehicles with enough
onboard computational capacity to run deliberative agents such as \rx
make water-column measurements and reliably track and characterize
dynamic features. While our experiments have already shown such
capabilities the exploration Vs. exploitation tradeoff is better
determined when vehicles can make choices over longer time periods
(weeks and months) to study the impacts of bio-geochemical interaction
driven by longer duration physical forcing.

Given our poor understanding of the coastal ocean in particular,
mixed-initiative methods of command and control for Sampling with
heterogenous assets is an important goal that is clearly at our
doorstep. We believe deliberation is small yet critical part of that
unfolding research problem for AI and Robotics.

### 2012-01-27 trunk/results.tex
updates

\paragraph {Lagrangian Surveys:} More recent experiments with \rx
controlling the Dorado AUV have centered on the \texttt{CANON}
(Controlled, Agile, and Novel Observing Network) program at MBARI. The
fundamental science goals of the program are related to detecting and
tracking of spatio-temporal coastal ocean features \cite{canon}. The
dynamic features require a robotic platform to move with it, return
water samples from across and within gradients and provide contextual
observations to other drifting instruments. All of these observational
strategies are tied to vehicle adaptation which is intrinsically tied
to deliberation and difficult to accomplish otherwise.

Various adaptation methodologies in \texttt{CANON} are still being
investigated; however what has become a critical goal for the program
has been to provide contextual environmental information about water
patches. The form and content of such 'patches' varies; for instance
initial attempts focused on Harmful Algal Blooms (HABs). They have
large spatial ($>$ 50 km$^2$) and temporal (days to weeks) extent and
are visible from space with visible coloring on the ocean surface. The
direct impact of HABs is the introduction of toxins into the marine
(and as a consequence human) food chain. In addition,
\cite{anderson00} and \cite{hoagland06} estimate that the average
economic impact resulting from HABs in the United States has been
$\sim$\$75 million from 1987 to 2000. However, the processes behind
bloom initiation, evolution and collapse is poorly understood in part
due to the complex interactions between microbial communities and the
surrounding environment. As a result, our capacity to assess the range
of potential future scenarios that might result from ocean temperature
changes, acidification, or nutrient shifts is highly limited.

\begin{figure}[htpb]
\centering
\includegraphics[width=0.5\textwidth]{figs/dta-app.pdf}
\caption{\small{Overall setup of CANON lagrangian experiments with a
    drifting instrument pack. Various combinations of computational
    methods can be included in the Drifter Tracking App which is a
    part of the Oceanographic Decision Support System (ODSS)
    \cite{das11}.}}
\label{fig:dta-setup}
\end{figure}

To augment our understanding of such microbial communities, the Dorado
with \rx has been utilized in a range of activities in a
quasi-lagrangian mode. As the water mass is being tracked with a
drifting instrument, the vehicle is expected to move around it. While
a number of different patterns have been tested in simulation, the
majority of at-sea deployments have used a square pattern ($3X3$
km\textsuperscript{2} or a $1X1$ km\textsuperscript{2})\footnote{Often
  post-hoc reconstruction of the data set for visualization drives how
  much adaptation the vehicle can be allowed to undertake.}. Plan
synthesis occurs frequently, but within the context of a drifting
point generated by a drifter. Fig. \ref{fig:dta-setup} shows the
general setup of our experiments which can be tailored with the use of
various computational methods within the \texttt{DTA} app which is
contained within the Oceanographic Decision Support System (ODSS; see
below) \cite{das11}. GPS data from the free-floating drifter is sent
via satellite and processed onshore to determine a projection (speed
and bearing) of the drifter. This data is sent to the AUV; at the
completion of the current survey, the vehicle receives this data as a
high-level goal which is decomposed and synthesized as a survey
plan. Currently, the vehicle completes the survey without any further
dynamic updates from the drifter. In order to minimize the range of
errors that the survey needs to encapsulate (data latency or surface
or sub-surface currents for those drifters which have a drogue) the
estimation is a projection in time and space of the drifters centroid
sent with this goal from the shore-side app.

@@

\begin{figure*}
\centering
\subfloat[]{\label{fig:sep10-ship}\includegraphics[width=0.38\textwidth]{figs/sep10-canon-cruise.png}}
\subfloat[]{\label{fig:sep10-transects}\includegraphics[width=0.35\textwidth]{figs/september-iterations.pdf}}
  \caption{\small{The offshore CANON ESP-drifter following cruise
      \cite{das11b}. Fig. \ref{fig:sep10-ship} shows the ESP drifter
      foreground with the R/V \emph{Western Flyer} with the image
      taken from the R/V \emph{Zephyr}. Fig \ref{fig:sep10-transects}
      shows the 86 km transect of the Dorado following the ESP drifter
      about 100 Nm from California shores.}}
\label{fig:gulper}
\end{figure*}

@@

Successive lagrangian experiments also followed a more innovative and
robotically choreographed ``ballet'' between two AUVs, the Dorado and
the LRAUV \cite{bellingham10} in October 2010 as part of another
series of \texttt{CANON} field experiments. The slower vehicle with a
fluorometer tracked a targeted chlorophyll patch seeded by a
GPS-enabled drifter, sending periodic updates of the patch centroid to
shore via the ODSS following a setup similar to
Fig.\ref{fig:dta-setup}, to the Dorado running \rx.

@@

\begin{figure*}
\centering
\subfloat[The ODSS Concept]{\label{fig:odss-concept}\includegraphics[width=0.5\textwidth]{figs/odss-concept.pdf}}\\
\subfloat[Current Implementation for \texttt{CANON}]{\label{fig:odss-frontend}\includegraphics[width=0.5\textwidth]{figs/dss-screenshot.png}}
\caption{\small The Oceanographic Decision Support System (ODSS) is a
  single portal for planning, visualization, situational awareness,
  collaboration and archival of information before, during and after
  field experiments. The ODSS has been used since mid-year 2010 for
  \texttt{CANON} field experiments.}
\label{fig:odss}
\end{figure*}

Large scale oceanographic field experiments akin to those targeted by
\can are usually logistically constrained driven largely by ship
schedules.  Once a feature set has been identified for scientific
observation and science goals have been collated, the appropriate
robotic assets need to be assigned and deployed for observation and
sampling. This asset allocation in turn depends on parameters like
sensor payload, the speed of the robot \situ in comparison to the
spatio-temporal scales of the evolution of the phenomena being
sampled, the number of assets in a class, the kinds of measurements
needed (e.g synoptic or close-up), logistical considerations for
deployments such as battery charge needed or distance/time to deploy,
impact to long-range deployment plans among others. All drive the
strategy to balance short and long-term scientific and engineering
needs. Our \can deliberations during the 2010 field season for example,
were driven by all of the above \cite{das11}. Often these strategies
were rebalanced during each decision-making cycle, especially in light
of opportunistic science goals. For example, gliders running at $0.5$
knots were often retargeted when surface support vessels noted the
apparent dispersal of the bloom to subsurface layers when HABs were a
\texttt{CANON} target. However, their retargeting was carefully
monitored to balance the long-term desire of obtaining time series
within fixed tracks and with the need to ensure these assets were not
exposed to strong currents. Powered AUVs were targeted to chase after
advecting blooms, but the decision on their deployment locations and
survey areas too close to the surf-zone were avoided. In
patch-tracking experiments, the deployment of drifters close to shore
and where depths of $20m$ or less were encountered, was avoided even
when contrary to scientific intent. The need to obtain high-quality
data for post-hoc analysis was also driven by the response time of
data returned by experiment partners.

\begin{figure*}
\centering
\subfloat[]{\label{fig:odss-mtg}\includegraphics[width=0.3\textwidth]{figs/october-canon-meeting.png}}
\subfloat[]{\label{fig:odss-virtual}\includegraphics[width=0.33\textwidth]{figs/virtual-drifter-tool.png}}
\subfloat[]{\label{fig:odss-ipad}\includegraphics[width=0.3\textwidth]{figs/photo-ipad-2.jpg}}
\caption{\small Fig. \ref{fig:odss-mtg} shows a 2010 \can meeting with the
  ODSS in use while Fig. \ref{fig:odss-ipad} shows its use on a portable
  device.}
xf\label{fig:odss2}
\end{figure*}

@@

Additionally, our work on ODSS is driven by a vision of how marine
field robotics will likely be conducted in the future and the
necessary computational infrastructure that will augment decision
making, onshore as well as on-board. We believe ultimately ODSS will
combine capabilities inferential and data gathering and analysis
capabilities and more uniquely, provide an event response capability
that could trigger the deployment of a robot and/or adapt its sampling
strategy, all from a scientist's desktop on shore. Data returned from
robotic assets at sea will be filtered for signals associated with
feature(s) of interest along with information on its contextual
environment. The association between the signal and a prospective
event will be undertaken through tagging by experts and machine
learning techniques used in recommender
systems\cite{Adomavicius05}. Adaptive sampling strategies will then
balance decision-making \situ with that done onshore with
computational models and human agents, the subject of a new NSF
project in which the authors are participating.

Operationally, the ODSS has become a software system that the
scientist consults during a \can field campaign. S/he consults
up-to-date model outputs, synoptic views of ocean color from remote
sensing data and weather conditions for the next $24$-hour planning
cycle and then queries the system to compute candidate sampling
locations. In the near future, the ODSS will consult its situational
awareness database, incorporate model outputs projecting these outputs
into the space-time region of interest to generates candidate
hypotheses with associated utility metrics. With the scientist's
guidance the automated planner would then determine the most viable
robotic option(s) to be dispatched (underwater, surface or aerial),
given available resources, distance of travel, required payload and
the environment. For instance, unpowered autonomous gliders would be
chosen for longer-duration missions with smaller payload requirements
in areas of moderate to weak ocean currents. The automated planner
would balance these criteria to compute a viable trajectory plan for
the selected vehicle. Vehicles with more controllability would be
given a specified area to survey and the path generated \situ. Goals
are communicated to the robot after analysis on shore for selective
targeting of features by the conglomerate of vehicles in the ocean
being tracked by the ODSS. Thus far, the ODSS has provided a simple
yet functional tool for scientists to use in the daily complex
decision making and the following experiment and survey design
needs. It has helped train scientists to think of expanding their
repertoire of tools for the scientific process and in the process it
enabled a mixed team of engineers and scientists to interact at the
functional level of the systems' capabilities.

>>>>

\paragraph {Lagrangian Surveys:} More recent experiments with \rx
controlling the Dorado AUV have centered on the \can (Controlled,
Agile, and Novel Observing Network) program at MBARI. The fundamental
science goals of the program are related to detecting and tracking of
spatio-temporal coastal ocean features \cite{canon}. The dynamic
features require a robotic platform to move with it, return water
samples from across and within gradients and provide contextual
observations to other drifting instruments. All of these observational
strategies are tied to vehicle adaptation which is intrinsically tied
to deliberation and difficult to accomplish otherwise.

Various adaptation methodologies in \can are still being
investigated; however what has become a critical goal for the program
has been to provide contextual environmental information about water
patches. The form and content of such 'patches' varies; for instance
initial attempts focused on Harmful Algal Blooms (HABs). They have
large spatial ($>$ 50 km$^2$) and temporal (days to weeks) extent and
are visible from space with visible coloring on the ocean surface. The
direct impact of HABs is the introduction of toxins into the marine
(and as a consequence human) food chain. In addition,
\cite{anderson00} and \cite{hoagland06} estimate that the average
economic impact resulting from HABs in the United States has been
$\sim$\$75 million from 1987 to 2000. However, the processes behind
bloom initiation, evolution and collapse is poorly understood in part
due to the complex interactions between microbial communities and the
surrounding environment. As a result, our capacity to assess the range
of potential future scenarios that might result from ocean temperature
changes, acidification, or nutrient shifts is highly limited.

\begin{figure}[htpb]
\centering
\includegraphics[width=0.5\textwidth]{figs/dta-app.pdf}
\caption{\small{Overall setup of \can lagrangian experiments with a
    drifting instrument pack. Various combinations of computational
    methods can be included in the Drifter Tracking App which is a
    part of the Oceanographic Decision Support System (\od)
    \cite{das11}.}}
\label{fig:dta-setup}
\end{figure}

To augment our understanding of such microbial communities, the Dorado
with \rx has been utilized in a range of activities in a
quasi-lagrangian mode. As the water mass is being tracked with a
drifting instrument, the vehicle is expected to move around it. While
a number of different patterns have been tested in simulation, the
majority of at-sea deployments have used a square pattern ($3X3$
km\textsuperscript{2} or a $1X1$ km\textsuperscript{2})\footnote{Often
  post-hoc reconstruction of the data set for visualization drives how
  much adaptation the vehicle can be allowed to undertake.}. Plan
synthesis occurs frequently, but within the context of a drifting
point generated by a drifter. Fig. \ref{fig:dta-setup} shows the
general setup of our experiments which can be tailored with the use of
various computational methods within the \texttt{DTA} app which is
contained within the Oceanographic Decision Support System (\od; see
below) \cite{das11}. GPS data from the free-floating drifter is sent
via satellite and processed onshore to determine a projection (speed
and bearing) of the drifter. This data is sent to the AUV; at the
completion of the current survey, the vehicle receives this data as a
high-level goal which is decomposed and synthesized as a survey
plan. Currently, the vehicle completes the survey without any further
dynamic updates from the drifter. In order to minimize the range of
errors that the survey needs to encapsulate (data latency or surface
or sub-surface currents for those drifters which have a drogue) the
estimation is a projection in time and space of the drifters centroid
sent with this goal from the shore-side app.

@@

\begin{figure*}
\centering
\subfloat[]{\label{fig:sep10-ship}\includegraphics[width=0.38\textwidth]{figs/sep10-canon-cruise.png}}
\subfloat[]{\label{fig:sep10-transects}\includegraphics[width=0.35\textwidth]{figs/september-iterations.pdf}}
  \caption{\small{The offshore \can ESP-drifter following cruise
      \cite{das11b}. Fig. \ref{fig:sep10-ship} shows the ESP drifter
      foreground with the R/V \emph{Western Flyer} with the image
      taken from the R/V \emph{Zephyr}. Fig \ref{fig:sep10-transects}
      shows the 86 km transect of the Dorado following the ESP drifter
      about 100 Nm from California shores.}}
\label{fig:gulper}
\end{figure*}

@@

Successive lagrangian experiments also followed a more innovative and
robotically choreographed ``ballet'' between two AUVs, the Dorado and
the LRAUV \cite{bellingham10} in October 2010 as part of another
series of \can field experiments. The slower vehicle with a
fluorometer tracked a targeted chlorophyll patch seeded by a
GPS-enabled drifter, sending periodic updates of the patch centroid to
shore via the \od following a setup similar to
Fig.\ref{fig:dta-setup}, to the Dorado running \rx.

@@

\begin{figure*}
\centering
\subfloat[The \od Concept]{\label{fig:odss-concept}\includegraphics[width=0.5\textwidth]{figs/odss-concept.pdf}}\\
\subfloat[Current Implementation for \can]{\label{fig:odss-frontend}\includegraphics[width=0.5\textwidth]{figs/dss-screenshot.png}}
\caption{\small The Oceanographic Decision Support System (\od) is a
  single portal for planning, visualization, situational awareness,
  collaboration and archival of information before, during and after
  field experiments. The \od has been used since mid-year 2010 for
  \can field experiments.}
\label{fig:odss}
\end{figure*}

Large scale oceanographic field experiments akin to those targeted by
\can are usually logistically constrained driven largely by ship
schedules.  Once a feature set has been identified for scientific
observation and science goals have been collated, the appropriate
robotic assets need to be assigned and deployed for observation and
sampling. This asset allocation in turn depends on parameters like
sensor payload, the speed of the robot \situ in comparison to the
spatio-temporal scales of the evolution of the phenomena being
sampled, the number of assets in a class, the kinds of measurements
needed (e.g synoptic or close-up), logistical considerations for
deployments such as battery charge needed or distance/time to deploy,
impact to long-range deployment plans among others. All drive the
strategy to balance short and long-term scientific and engineering
needs. Our \can deliberations during the 2010 field season for example,
were driven by all of the above \cite{das11}. Often these strategies
were rebalanced during each decision-making cycle, especially in light
of opportunistic science goals. For example, gliders running at $0.5$
knots were often retargeted when surface support vessels noted the
apparent dispersal of the bloom to subsurface layers when HABs were a
\can target. However, their retargeting was carefully
monitored to balance the long-term desire of obtaining time series
within fixed tracks and with the need to ensure these assets were not
exposed to strong currents. Powered AUVs were targeted to chase after
advecting blooms, but the decision on their deployment locations and
survey areas too close to the surf-zone were avoided. In
patch-tracking experiments, the deployment of drifters close to shore
and where depths of $20m$ or less were encountered, was avoided even
when contrary to scientific intent. The need to obtain high-quality
data for post-hoc analysis was also driven by the response time of
data returned by experiment partners.

\begin{figure*}
\centering
\subfloat[]{\label{fig:odss-mtg}\includegraphics[width=0.3\textwidth]{figs/october-canon-meeting.png}}
\subfloat[]{\label{fig:odss-virtual}\includegraphics[width=0.33\textwidth]{figs/virtual-drifter-tool.png}}
\subfloat[]{\label{fig:odss-ipad}\includegraphics[width=0.3\textwidth]{figs/photo-ipad-2.jpg}}
\caption{\small Fig. \ref{fig:odss-mtg} shows a 2010 \can meeting with the
  \od in use while Fig. \ref{fig:odss-ipad} shows its use on a portable
  device.}
xf\label{fig:odss2}
\end{figure*}

@@

Additionally, our work on \od is driven by a vision of how marine
field robotics will likely be conducted in the future and the
necessary computational infrastructure that will augment decision
making, onshore as well as on-board. We believe ultimately \od will
combine capabilities inferential and data gathering and analysis
capabilities and more uniquely, provide an event response capability
that could trigger the deployment of a robot and/or adapt its sampling
strategy, all from a scientist's desktop on shore. Data returned from
robotic assets at sea will be filtered for signals associated with
feature(s) of interest along with information on its contextual
environment. The association between the signal and a prospective
event will be undertaken through tagging by experts and machine
learning techniques used in recommender
systems\cite{Adomavicius05}. Adaptive sampling strategies will then
balance decision-making \situ with that done onshore with
computational models and human agents, the subject of a new NSF
project in which the authors are participating.

Operationally, the \od has become a software system that the
scientist consults during a \can field campaign. S/he consults
up-to-date model outputs, synoptic views of ocean color from remote
sensing data and weather conditions for the next $24$-hour planning
cycle and then queries the system to compute candidate sampling
locations. In the near future, the \od will consult its situational
awareness database, incorporate model outputs projecting these outputs
into the space-time region of interest to generates candidate
hypotheses with associated utility metrics. With the scientist's
guidance the automated planner would then determine the most viable
robotic option(s) to be dispatched (underwater, surface or aerial),
given available resources, distance of travel, required payload and
the environment. For instance, unpowered autonomous gliders would be
chosen for longer-duration missions with smaller payload requirements
in areas of moderate to weak ocean currents. The automated planner
would balance these criteria to compute a viable trajectory plan for
the selected vehicle. Vehicles with more controllability would be
given a specified area to survey and the path generated \situ. Goals
are communicated to the robot after analysis on shore for selective
targeting of features by the conglomerate of vehicles in the ocean
being tracked by the \od. Thus far, the \od has provided a simple
yet functional tool for scientists to use in the daily complex
decision making and the following experiment and survey design
needs. It has helped train scientists to think of expanding their
repertoire of tools for the scientific process and in the process it
enabled a mixed team of engineers and scientists to interact at the
functional level of the systems' capabilities.
